---
title: "COViD project analysis"
author: 
- "Szymon M. Kiełbasa (BDS/LUMC)"
- "Mo Arkani (BDS/LUMC)"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    code_folding: hide
---

# COViD project analysis

```{r echo=FALSE, warning=FALSE, message=FALSE}
source( "covid.R" )
covData <- CovidData$new( 
  "Dataset/BREAK COVID overview v3_combined_data__20240310_LUMC.xlsx",
  vaxValidDaysRange = c( 14, 270 ), reinfDaysRange = c( 14, 270 )
)
immData <- ImmunData$new( "Dataset/IMMUN_COVID20240130_V2.xlsx" )
dm <- DataMerger$new( 
  immData, covData, transDaysRange = c( 14, 14+9*31 ), preferredDaysDiff = 180,
  projectSources = c( "CoViCross", "CBT" ) # select one or more sources
)
mm <- dm$getMeasures()
dp <- DataProcessor$new( dm )
```

## Samples {.tabset .tabset-fade}

### Collection (sample)

```{r fig.width=12, fig.height=5}
set.seed( 124L )
selSubjectIds <- dp$getSubjectIds() |> sample( 30L ) |> sort()
dp$eventsPlot( showTransitions = FALSE, subjects = selSubjectIds )
```

Each row denotes **one subject** (total: `r length( dp$getSubjectIds() )`).  
Each letter denotes **one blood sample** and subject state: **control** (B or C), **case** (A - asymptomatic; S - symptomatic), unknown (D - dubious).  
The blue stars denote **vaxination** moments. Within 9 months break-through-vaccination (blue) and reinfection (red) events are highlighted.  
Different markers were measured at different timepoints and for different subjects (there is no full symmetry; e.g. squares denote the only samples which were chosen for flow cytometry analysis).

### Pairs (sample)

```{r fig.width=12, fig.height=5}
dp$eventsPlot( showTransitions = TRUE, subjects = selSubjectIds )
```

Pairs of measurements at **two timepoints** (transitions from T1  to T2) were created for further analysis.  
The pairs were selected for each subject, and for each marker, separately.  
The pairs were classified into two groups: transitions from **control to case** (red), and transitions from **control to control** (green).  
They were chosen to maximize the number of control-case pairs, and to prefer approx. 6 months between the timepoints.

### Collection (all)
```{r fig.width=12, fig.height=25}
dp$eventsPlot( showTransitions = FALSE )
```

### Transitions (all)

```{r fig.width=12, fig.height=25}
dp$eventsPlot( showTransitions = TRUE )
```

## Markers and possible analyses {.tabset .tabset-fade}

### List

Serology dataset:

```{r comment=NA}
cat( paste0( covData$getMarkers(), collapse = ", " ) )
```

Cytometry dataset:

```{r comment=NA}
cat( paste0( immData$getMarkers(), collapse = ", " ) )
```

### Level and state

```{r fig.width=12, fig.height=5}
selMarkers <- c( "BA1", "NCP", "T_cells" )
dp$markersTimePlots( showTransitions = FALSE, showMeasurements = "events", markers = selMarkers )
```

A sample of measured markers and their levels at each measured timepoint are shown.  
The point color represents the **current state** of the subject (control - green, asymptomatic case - orange, symptomatic case - red).

### Level change (T1->T2) and state (T2)

```{r fig.width=12, fig.height=5}
dp$markersTimePlots( showTransitions = TRUE, showMeasurements = NULL, markers = selMarkers )
```

**Changes** of marker levels between current and later moment are shown.  
Transitions are split into two groups: **control->case** (red) vs. **control->control** (green).

### Level (T1) to prognose state (T2)

```{r fig.width=12, fig.height=5}
dp$markersTimePlots( showTransitions = TRUE, transitionAlpha = 0.1, showMeasurements = "T1events", markers = selMarkers )
```

The central question: is it possible to **predict** whether a subject **will become a case or control** in the future, based on the marker levels at the current timepoint?

## Level and state {.tabset .tabset-fade}

### Serology markers

```{r fig.width=12, fig.height=7}
d <- dp$calcMeasurementsDataWilcoxTest( formula = Value ~ State, markers = covData$getMarkers() )
dp$markersTimePlots( showTransitions = FALSE, showMeasurements = "events", markers = d$Marker )
```

### ←Stats

Wilcoxon test for difference in markers between cases and controls.
```{r results='asis', warning=FALSE}
knitr::kable( d )
```

### ←Boxplot
```{r fig.width=12, fig.height=7}
dp$markersBoxPlots( markers = d$Marker )
```

### Cytometry markers

```{r fig.width=12, fig.height=7}
d <- dp$calcMeasurementsDataWilcoxTest( formula = Value ~ State, markers = immData$getMarkers() )
dp$markersTimePlots( showTransitions = FALSE, showMeasurements = "events", markers = d$Marker )
```

### ←Stats

Wilcoxon test for difference in markers between cases and controls.

```{r results='asis', warning=FALSE}
knitr::kable( d )
```

### ←Boxplot
```{r fig.width=12, fig.height=7}
dp$markersBoxPlots( markers = d$Marker )
```

## Level change (T1->T2) and state (T2) {.tabset .tabset-fade}

### Serology markers

```{r fig.width=12, fig.height=7}
d <- dp$calcTransitionDataWilcoxTest( formula = Value - prevValue ~ TransitionName, markers = covData$getMarkers() )
dp$markersTimePlots( showTransitions = TRUE, showMeasurements = NULL, markers = d$Marker )
```

### ←Stats

Wilcoxon test for marker difference (T2-T1) dependence on endpoint case/control (at T2).

```{r results='asis', warning=FALSE}
knitr::kable( d )
```

### ←Scatterplot

```{r fig.width=12, fig.height=7}
dp$markersT1T2ScatterPlots( markers = d$Marker )
```

### Cytometry markers

```{r fig.width=12, fig.height=7}
d <- dp$calcTransitionDataWilcoxTest( formula = Value - prevValue ~ TransitionName, markers = immData$getMarkers() )
dp$markersTimePlots( showTransitions = TRUE, showMeasurements = NULL, markers = d$Marker )
```

### ←Stats

Wilcoxon test for marker difference (T2-T1) dependence on endpoint case/control (at T2).

```{r results='asis', warning=FALSE}
knitr::kable( d )
```

### ←Scatterplot

```{r fig.width=12, fig.height=7}
dp$markersT1T2ScatterPlots( markers = d$Marker )
```

## Level (T1) to prognose state (T2) {.tabset .tabset-fade}

### Serology markers

```{r fig.width=12, fig.height=7}
d <- dp$calcTransitionDataWilcoxTest( formula = prevValue ~ TransitionName, markers = covData$getMarkers() )
dp$markersTimePlots( showTransitions = TRUE, transitionAlpha = 0.1, showMeasurements = "T1events", markers = d$Marker )
```

### ←Stats (Wilcoxon)

Wilcoxon test for marker (at T1) which could predict future **case/control** (at T2).

```{r results='asis', warning=FALSE}
knitr::kable( d )
```

### ←Stats (Anova/TukeyHSD)

Anova/TukeyHSD test for marker (at T1) which could predict future **symptomatic or asymptomatic** case/control (at T2).

```{r}
l <- dp$calcTransitionDataKruskalTest( markers = covData$getMarkers() )
knitr::kable( l$kruskalTestsD )
```

```{r}
knitr::kable( l$wilcoxTestsD )
```

### ←Boxplot

```{r fig.width=12, fig.height=7}
dp$markersT1forTransitionsBoxPlots( markers = d$Marker )
```

### Cytometry markers

```{r fig.width=12, fig.height=7}
d <- dp$calcTransitionDataWilcoxTest( formula = prevValue ~ TransitionName, markers = immData$getMarkers() )
dp$markersTimePlots( showTransitions = TRUE, transitionAlpha = 0.1, showMeasurements = "T1events", markers = d$Marker )
```

### ←Stats

Wilcoxon test for marker (at T1) which could predict future case/control (at T2).

```{r results='asis', warning=FALSE}
knitr::kable( d )
```

### ←Boxplot

```{r fig.width=12, fig.height=7}
dp$markersT1forTransitionsBoxPlots( markers = d$Marker )
```

## Add-ons {.tabset .tabset-fade}

### Age/gender

Only timewise-first measurements of each subject are considered. Red line denotes the average age.

```{r}
dp$ageT1DistPlot()
```

### Markers (VBT/reinf)

```{r fig.width=12, fig.height=11}
dp$markersTimePlots( showTransitions = TRUE, showMeasurements = "vaxReInf" )
```

### Boxplot (VBT)

Markers at T1 for subjects classifed VBT at T2.

```{r fig.width=12, fig.height=11}
dp$vaxReinfT1BoxPlots( "vax" )
```

### Boxplot (reinf)

Markers at T1 for subjects classifed reinf at T2.

```{r fig.width=12, fig.height=11}
dp$vaxReinfT1BoxPlots( "reinf" )
```

## Document info

```{r echo=FALSE}
repo <- git2r::repository( "." )
l <- list()
l$sha <- git2r::last_commit(repo)$sha
l$url <- paste0( "https://", gsub( "[.]git$", "", gsub( ":", "/", gsub( "^[^@]*[@]", "", git2r::remote_url(repo) ) ) ), "/-/tree/", l$sha )
l$stat <- lapply( git2r::status( repo ), length )
l$stat <- l$stat[ l$stat > 0 ]
l$state <- if( do.call( sum, l$stat ) > 0 ) paste0( " [", paste( names( l$stat ), l$stat, sep = ":", collapse = ", " ), "]" ) else ""
```

Last commit: [`r paste0( l$sha, l$state )`](`r l$url`).